\scuba: iterative map-making with the Sub-Millimetre User Reduction Facility

### 2012-09-12 smurf_paper.tex
more edits from Douglas

There are numerous ways to attack this map-making problem, both in
terms of the data-collection method, and processing. The two most
important principles to follow in terms of scan strategy are: (i) to
modulate the astronomical signals of interest in such a way that they
appear in the lowest-noise regions of the bolometer noise power
spectrum, i.e., above the $1/f$ knee; and (ii) to provide good
``cross-linking'', in which each portion of the map is sampled on a
range of temporal scales, again, to help distinguish time-varying
noise features from fixed astronomical sources. In the case of \scuba,
(i) is achieved through fast-scanning of the entire telescope (up to
600\,arcsec\,sec$^{-1}$ slew speeds), such that significant drift in
the bolometers due to low-frequency noise occurs more slowly than the
crossing times for the astronomical scales of interest; and (ii) by
offering scan patterns that cross the sky at a wide range of position
angles, ultimately observing each portion of the map at a number of
different times. Such methods are now used by virtually all existing
ground-based submillimetre cameras
\citep[e.g.,][]{glenn1998,weferling2002,wilson2008,kovacs2008b}, in
preference to ``chopping'' methods (where the secondary is quickly
moved to modulate the signal) that were more appropriate for older
instruments that had poorer low-frequency noise performance, and were
only sensitive to modest angular scales

@@

Incoming light passes through a beam-splitter, and then bandpass
filters, providing simultaneous illumination and wavelength definition
of both the 450 and 850\,\micron\ focal planes. Each focal plan is
populated by four ``subarrays'' (labeled s4a--s4d at 450\,\micron, and
s8a--s8d at 850\,\micron), each consisting of 32 columns and 40 rows
of bolometers. The bolometers are thermal absorbers coupled to
superconducting transition-edge sensors (TESs) for thermometry.
Temperature variations in the TESs produce changing currents, and
therefore varying magnetic fields, which are detected and amplified
using chains of superconducting quantum interference devices (SQUIDs),
before the larger output currents are digitised. Each detector has its
own SQUID for the first-stage of the amplification, but the remaining
stages occur within a common chain of SQUIDs for each of the 32
columns. All 40 rows are read out in sequence at a row-visit frequency
of about 12\,kHz. Such a high sample rate is unnecessary to produce
maps, so the data are re-sampled to approximately 200\,Hz before
writing to disk. This rate provides a sample every 3.0\,arcsec, or
approximately one third of the 450\,\micron\ diffraction-limited
full-width at half-maximum (FWHM) -- a typical rule-of-thumb for
adequately sampling a Gaussian point spread function (PSF). There is
an additional 41st row of SQUID readouts that are not connected to
TESs. These ``dark SQUIDs'' track non-thermal noise sources that are
common to each column's amplifier chain. The relationship between the
output digitized current, $I$, and the input power, $P$, or $dI/dP$ is
established using flatfield observations immediately prior to science
observations, in which the output signal is measured throughout a ramp
of the pixel heaters (which provide a known input power); see
Section~2.1 in \citet{dempsey2012} and Section~5.3 in
\citet{holland2012} for more details. Finally, the conversion to
astronomical flux units from power involves a correction for
atmospheric extinction \citep[primarily using the JCMT Water Vapour
Monitor to track line-of-sight opacity variations, see Section~3
in][]{dempsey2012}, and the application of a flux conversion factor
(FCF) which is established from regular measurements of calibrators
such as Uranus \citep[Section~5 in][]{dempsey2012}.

@@

Components~6 and 8--10 in Fig.~\ref{fig:pca} have several generic
features in common. First, the PSDs exhibit moderate low-frequency
$1/f$ noise. Second, there are broad-line features at frequencies
above $\sim$30\,Hz which we believe to be aliased high-frequency
noise, in part due to the clear correlation patterns that appear along
columns in the eigenvalue maps at both wavelengths (e.g., columns 26
and 29 at 450\,\micron in Component~6). Finally, these components have
potentially unrelated features mixed into them, like sudden offsets
that are likely residuals due to imperfect step correction (e.g., at
$t\sim50$\,s in the eigenvector time-series for Component~6; also see
Section~\ref{sec:steps}).

Finally, Component~12 in Fig.~\ref{fig:pca} is significant in only a
handful of detectors in the 850\,\micron\ subarray (e.g., column 21
row 25). The time-series plot for the eigenvector exhibits a large
positive tail of sudden excursions which gradually worsen over
time. Bolometers that exhibit this signal are probably poorly biased,
and are usually flagged due to elevated noise levels during
pre-processing (Section~\ref{sec:flagbadbol}).

This example illustrates some of the types of correlated signals and
patterns that PCA can identify in \scuba\ data. While there are
typically one or two strong signal components detected, in general,
the details can vary significantly from data set to data set, and the
lengths of the time-series analysed. The reason for this is probably
due to the fact that many of the noise sources are not stationary in
time (e.g., scan-synchronous noise which obviously depends on the scan
pattern, and the tuning of the subarrays). It should also be clear
from this example that while PCA offers some helpful insight into the
various sources of noise, it does not necessarily identify clean
patterns that can easily be modelled. For example, while there are
clearly correlation patterns along columns (Components~6 and 8--10,
particularly at 450\,\micron), as one might expect given the common
amplification chain, the intensities of these high-frequency signals
are random in the eigenvalue maps, and a common-mode signal estimated
from the data for each column would not remove it. Finally, we note
that most of the high-frequency ($\gsim2$\,Hz) correlated features
that have been identified tend to be restricted to a small number of
detectors, and can often be identified and removed using other
(faster) means. What remains is predominantly low-frequency
noise. This conclusion is central to our data reduction strategy
described in the next section.

@@

The approach taken by SMURF to reducing \scuba\ data is to model (and
remove) predominantly low-frequency noise sources that are correlated
amongst detectors, and to iterate this process along with estimates of
the map. Significant experimentation with different models for noise
sources during \scuba\ commissioning required a highly flexible
software framework, and a configurable user interface. To achieve this
goal, while minimizing development time, it was decided to build SMURF
as a Starlink package \citep{2009ASPC..411..418J}, which provides
access to a large suite of libraries (including a commanding and
messaging interface, astrometric coordinate conversions and generation
of standard WCS information, file formats etc.). Furthermore,
Starlink\footnote{\url{http://www.starlink.ac.uk}} is
open-source\footnote{\url{https://github.com/Starlink}}, and already
used extensively at the Joint Astronomy Centre (host of the JCMT) for
many other systems \citep{jenness2011}, which helps with
interoperability. Though originally written in FORTRAN, many of the
core Starlink libraries are now ported to native C, or at least have a
C interface. It was therefore decided to develop SMURF in C as well
(rather than C++, for example, which would have required adding
further dependencies to Starlink), although we have taken an
object-oriented approach. For example, all of the data for a given
noise model is encapsulated in a C structure, and there is a standard
interface for all functions that handle models (member data and
functions for the class, respectively). In this way it is easy to
extend SMURF with new models. Parallelisation is incorporated in the
most time-consuming low-level routines using threads (e.g., performing
Fast Fourier Transforms, re-gridding the data), usually handling
either independent blocks of bolometers, or blocks of time, in each
thread, depending on the nature of the calculation. We have found that
for typical data sets the processing time scales well with the number
of central processing unit (CPU) cores, although beyond 8 the returns
are diminished due to tasks that cannot be parallelised (e.g., reading
the data from disk).

@@

First, the raw data are passed through a pre-processing stage which
corrects some of the more significant glitches, applies flat-field
corrections etc. Next, the iterative process begins (dashed box). Most
of the low-frequency noise is removed using common-mode subtraction
(\model{COM,GAI}, Section~\ref{sec:comgai}). Then, the extinction
correction (\model{EXT}, Section~\ref{sec:ext}) is applied. At this
stage the data resemble the grey bolometer traces in
Fig.~\ref{fig:bolos_mix}, with some residual baseline drifts still
visible. These drifts are removed using a high-pass filter
(\model{FLT}, Section~\ref{sec:flt}) implemented with Fast Fourier
Transforms (FFTs). The residual signal is then re-gridded to estimate
the map. Finally, the map is used to estimate and remove the
astronomical signal contamination in the bolometer data (\model{AST},
Section~\ref{sec:ast}), leaving a relatively clean data set in which
to measure the white noise level of each bolometer (\model{NOI},
Section~\ref{sec:noi}). Since the common-mode and filtering stages
will have introduced ringing in the map in the vicinity of bright
sources, the entire process is iterated. Each component in the dashed
box of Fig.~\ref{fig:dimm} is re-calculated in sequence. In this way,
the second time the common-mode is calculated, for example, most of
the bright astronomical sources in the data have already been removed
in the previous iteration when the map was estimated, reducing the
amount of ringing in the map once it is re-estimated. Note that the
extinction correction is a multiplicative factor, and must be inverted
prior to re-calculating any of the additive model components to
preserve the data amplitude. Additive model components from the
previous iteration are generically added back into the time-series
immediately prior to their re-calculation. However, for the special
case of the common-mode and high-pass filter stages, they are replaced
simultaneously at the start of the iteration to assist with
convergence (see Sections~\ref{sec:converge}).

@@

The method used by SMURF is to average together multiple samples from
the original time-series, $x_i$ to estimate the lower-frequency output
time-series, $y_j$. In general there will not be an integer number of
samples from $x_i$ in each of the $y_j$, so fractional samples from
$x_i$ are used at the $y_j$ time boundaries. The algorithm is fast,
since each sample in $x_i$ need only be visited once. From a spectral
point of view, this boxcar average is equivalent to applying a
sinc-function low-pass filter. Such behaviour is desirable since it
serves as an anti-aliasing filter; most non-white features above the
Nyquist frequency in $y_j$ that could contaminate the output data are
removed.

@@

In addition to replacing bad samples before the FFT, it is also
necessary to ensure that the data values at the start and end of the
time-series are similar. Since an FFT treats the data as a single
cycle in an infinitely repeating waveform, any large difference
between starting and ending value will effectively introduce sudden
steps at the start and end of each cycle, causing unwanted
oscillations (ringing) in the transform. Another consequence of the
cyclic nature of the FFT is that features at one end of the time
series can affect the filtered values at the other end of the time
series. Two methods are available to avoid these two problems:

@@

\item Apodisation: a number of samples at the start and end of each
bolometer time-series are multiplied by a cosine function in order to
roll the data values off smoothly to zero. The default number of
samples modified at each end of the time-series is given by half the
ratio of the sampling frequency to the lowest retained frequency. In
addition, each end of the time-series is padded with double this
number of zeros. This method is illustrated in
Fig.~\ref{fig:pad2}. This method is not used by default as it reduces
the amount of data available for the map, and can significantly hinder
very short observations (e.g., of calibrators when focussing the
telescope).

@@

Despite the cleaning operations described in the previous sections,
the data from a given bolometer may be unusable do to it being poorly
biased, having an incorrect flatfield correction applied, or having
some other particularly pathological noise contamination. Most of
these bolometers can be flagged simply by identifying outliers in the
distribution of bolometer white noise levels. By default, SMURF
measures the PSD of all bolometers between 2 and 10\,Hz, after all
other pre-processing steps have been run (identical to the measurement
in Section~\ref{sec:noi}). Despite having significant low-frequency
noise (and possibly bright astronomical source) contamination, this
higher-frequency portion of the PSD is generally quite
clean. Furthermore, even if there is contamination, the purpose of
this measurement is to identify outliers, rather than provide a
meaningful absolute measurement of a given bolometer's white noise
level. Both high and low (e.g., due to an incorrect and very small
flatfield correction being applied) outliers from the centre of the
logarithm (to reduce the impact of outliers) of the bolometer noise
distribution are flagged.

@@

We then express the noise, $\mathbf{n}_i(t)$, as the sum of several
components,
%
\begin{equation}
  \mathbf{n}_i(t) = \mathbf{n}^\mathrm{w}_i(t) +
  g_i\mathbf{n}^\mathrm{c}(t) + \mathbf{n}^\mathrm{f}_i(t),
\label{eq:noise}
\end{equation}
%
where $\mathbf{n}^\mathrm{w}_i(t)$ is uncorrelated white noise,
$\mathbf{n}^\mathrm{c}(t)$ is a correlated or common-mode signal (with
an optional scale factor $g_i$ for each bolometer), and
$\mathbf{n}^\mathrm{f}_i(t)$ is (predominantly low-frequency) noise in
excess of the white noise level that is either uncorrelated from
bolometer-to-bolometer, or at least does not have a simple correlation
relationship that would lead to it being included in
$\mathbf{n}^\mathrm{c}(t)$.

@@

\begin{table}
  \caption{Summary of the model components that can be fit to
    \scuba\ time-series data with SMURF. Only the first group of
    models (in the order indicated) are typically fit to the data
    (\model{COM}--\model{NOI}) in the indicated order. The remaining
    models (\model{DKS}--\model{PLN}) have only been included for
    completeness}
  \vspace{0.2cm}
  \centering
  \begin{tabular}{c|l}
    \hline
    Model & Description \\
    \hline
    \model{COM} & remove common-mode signal \\
    \model{GAI} & common-mode scaled to each bolometer \\
    \model{EXT} & extinction correction \\
    \model{FLT} & Fourier transform filter \\
    \model{AST} & map estimate of astronomical signal \\
    \model{NOI} & noise estimation \\
    \hline
    \model{DKS} & dark squid cleaning along columns \\
    \model{PLN} & 2-dimensional time-varying plane removal \\
    \model{SMO} & time-domain smoothing filter \\
    \model{TMP} & pointing as baseline template \\
    \hline
    \end{tabular}
  \label{tab:components}
\end{table}

@@

\begin{figure}
\centering
\includegraphics[width=\linewidth]{com.pdf}
\caption{A selection of typical bolometer time-series after
flatfielding and removal of a constant baseline. It can be seen that
most bolometers exhibit a common time variation overlayed with other
features. Empty squares indicate the locations of dead bolometers.}
\label{fig:com}
\end{figure}

The \model{COM} model is the common-mode signal itself
[$\mathbf{n}^{c}(t)$ in Eq.~\ref{eq:noise}]. It is a single time
series that is estimated by finding the mean of all bolometers at each
time step. Bolometer values that have been flagged as unusable are
excluded from the mean.

@@

It can be seen from Fig.~\ref{fig:com} that some bolometers depart
radically from the common-mode, indicating some problem with the
bolometer. Such bolometers are identified by calculating the Pearson
coefficient of correlation between each bolometer time-series and the
common mode. Bolometers for which the correlation coefficient is below
a specified limit, or which have unusually high or low gains compared
to the other bolometers, are flagged as bad in order to omit them from
the final map. If the option described above for handling time-varying
sensitivities is used, then a correlation coefficient can be
determined for each individual block of time slices.  This allows
individual bad blocks to be rejected from a bolometer time-series,
rather than rejecting the whole bolometer. As a warning, rejecting
data based on outlier gain values can be misleading in cases where the
data are dominated by magnetic field pickup. For example, the
common-mode signal for the data in Fig.~\ref{fig:magpickup} would
resemble the azimuthal scan pattern, but clearly the gains will have
opposite signs (such that one or the other bolometer would be
erroneously flagged).

@@

Any extended astronomical emission on a scale comparable to or larger
than the spatial extent of the area used to estimate the common-mode
will contribute a similar signal to all bolometers. Therefore such
extended emission is indistinguishable from the other sources of
common-mode signal (e.g., atmosphere variations) and will be removed by
the model. This places a limit on the spatial extent of astronomical
structure that can be recovered.

@@

This model takes the FFT of the bolometer time-series data, and can
apply both high- and low-pass filters, as well as notch filteres, at
hard frequency edges specified by the user. Alternatively, the
frequency edges of the filters may be defined in terms of an angular
scale, but converted into a frequency through knowledge of the mean
telescope slew speed. The time-series are generally gap-filled
(Section~\ref{sec:gaps}) before the transform to avoid ringing
(primarily caused by wrap-around discontinuities at the ends of the
time-series). Finally, a whitening filter may also be applied in which
a simple $1/f^\alpha + \mathrm{constant}$ is fit to the power spectrum
of each bolometer, and then the bolometer FFT is multiplied by the
square root of its inverse (though normalised to preserve the white
noise level).  The signals that are removed from the time-series by
this process are stored in the \model{FLT} container array. Typically
this model is used purely as a high-pass filter to remove most of the
residual $1/f$ noise following common-mode removal
[$\mathbf{n}^\mathrm{f}_i(t)$ in Eq.~\ref{eq:noise}]. Low-pass
filtering is redundant for two reasons: (i) SMURF already low-pass
filters and re-samples to a lower sample rate, as described in
Section~\ref{sec:downsamp}; and (ii) the act of re-gridding the data
to produce map estimates effectively low-pass filters the data below a
frequency that corresponds to the inverse of the crossing time of a
single map pixel. Notch filters have not been proven to be useful with
\scuba\ data, particularly since line features tend to move around
(i.e., a dynamic line-detection system would have to be developed, and
care would have to be taken that astronomical sources are not
suppressed).

@@

In addition to the signal map, a variance map $\mathbf{v}(x_j,y_j)$ is
estimated. The default behaviour is to estimate this quantity given
the scatter in the weighted samples. This is accomplished by dividing
the biased weighted sample variance by the number of samples that went
into the average (akin to the formula for standard error on the mean,
but accounting for weights),
%
\begin{equation}
\label{eq:varmap}
\mathbf{v}(x_j,y_j) = \frac{\sum_j \mathbf{w}_j
                            \sum_j \mathbf{w}_j \mathbf{b}_j -
                            \left( \sum_j \mathbf{w}_j \mathbf{b}_j \right)^2 }
                           { N \left( \sum_j \mathbf{w}_j \right)^2 },
\end{equation}
%
where $N_j$ is the total number of bolometer samples that land in the
pixel. As written, this algorithm is numerically unstable if the two
terms in the numerator are large, and of nearly the same value:
floating point precision errors can cause the difference to be
significantly incorrect. In practice we use the more numerically
stable ``weighted incremental algorithm'' that calculates incremental
differences, as described in \citet{west1979}.

We decided not to use an unbiased estimator (e.g., the extension of
the common $(1/N-1)\sum_j (\mathbf{b}_j - \bar{\mathbf{b}})^2$
estimator using weights), since in practice it would require
accumulating an additional array of values at every map pixel, and
only results in a small difference where there are less than $\sim10$
samples per pixel (a situation that is almost never encountered in a
\scuba\ map except in the edge pixels).

>>>>

There are numerous ways to attack this map-making problem, both in
terms of the data-collection method, and processing. The two most
important principles to follow in terms of scan strategy are: (i) to
modulate the astronomical signals of interest in such a way that they
appear in the lowest-noise regions of the bolometer noise power
spectrum, i.e., above the $1/f$ knee; and (ii) to provide good
``cross-linking'', in which each portion of the map is sampled on a
range of temporal scales, again, to help distinguish time-varying
noise features from fixed astronomical sources. In the case of \scuba,
(i) is achieved through fast-scanning of the entire telescope (up to
600\,arcsec\,sec$^{-1}$), such that significant drift in the
bolometers due to low-frequency noise occurs more slowly than the
crossing times for the astronomical scales of interest; and (ii) by
offering scan patterns that cross the sky at a wide range of position
angles, ultimately observing each portion of the map at a number of
different times. Such methods are now used by virtually all existing
ground-based submillimetre cameras
\citep[e.g.,][]{glenn1998,weferling2002,wilson2008,kovacs2008b}, in
preference to ``chopping'' methods (where the secondary is moved
quickly to modulate the signal) that were more appropriate for older
instruments that had poorer low-frequency noise performance, and were
only sensitive to modest angular scales

@@

Incoming light passes through a beam-splitter, and then bandpass
filters, providing simultaneous illumination and wavelength definition
of both the 450 and 850\,\micron\ focal planes. Each focal plan is
populated by four ``subarrays'' (labeled s4a--s4d at 450\,\micron, and
s8a--s8d at 850\,\micron), each consisting of 32 columns and 40 rows
of bolometers. The bolometers are thermal absorbers coupled to
superconducting transition-edge sensors (TESs) for thermometry.
Temperature variations in the TESs produce changing currents, and
therefore varying magnetic fields, which are detected and amplified
using chains of superconducting quantum interference devices (SQUIDs),
before the larger output currents are digitised. Each detector has its
own SQUID for the first-stage of the amplification, but the remaining
stages occur within a common chain of SQUIDs for each of the 32
columns. All 40 rows are read out in sequence at a row-visit frequency
of about 12\,kHz. Such a high sample rate is unnecessary to produce
maps, so the data are re-sampled to approximately 200\,Hz before
writing to disk. This rate provides a sample every 3.0\,arcsec, or
approximately one third of the 450\,\micron\ diffraction-limited
full-width at half-maximum (FWHM) -- a typical rule-of-thumb for
adequately sampling a Gaussian point spread function (PSF). There is
an additional 41st row of SQUID readouts that are not connected to
TESs. These ``dark SQUIDs'' track non-thermal noise sources that are
common to each column's amplifier chain. The relationship between the
output digitised current, $I$, and the input power, $P$, or $dI/dP$ is
established using flatfield observations immediately prior to science
observations, in which the output signal is measured throughout a ramp
of the pixel heaters (which provide a known input power); see
Section~2.1 in \citet{dempsey2012} and Section~5.3 in
\citet{holland2012} for more details. Finally, the conversion to
astronomical flux units from power involves a correction for
atmospheric extinction \citep[primarily using the JCMT Water Vapour
Monitor to track line-of-sight opacity variations, see Section~3
in][]{dempsey2012}, and the application of a flux conversion factor
(FCF) which is established from regular measurements of calibrators
such as Uranus \citep[Section~5 in][]{dempsey2012}.

@@

Components~6 and 8--10 in Fig.~\ref{fig:pca} have several generic
features in common. First, the PSDs exhibit moderate low-frequency
$1/f$ noise. Second, there are broad-line features at frequencies
above $\sim$30\,Hz which we believe to be aliased high-frequency
noise, in part due to the clear correlation patterns that appear along
columns in the eigenvalue maps at both wavelengths (e.g., columns 26
and 29 at 450\,\micron\ in Component~6). Finally, these components
have potentially unrelated features mixed into them, like sudden
offsets that are likely residuals due to imperfect step correction
(e.g., around $t\sim50$\,s in the eigenvector time-series for
Component~6; also see Section~\ref{sec:steps}).

Finally, Component~12 in Fig.~\ref{fig:pca} is significant in only a
handful of detectors in the 850\,\micron\ subarray (e.g., column 21
row 25). The time-series plot for the eigenvector exhibits a large
positive tail of sudden excursions, which gradually worsen over
time. Bolometers that exhibit this signal are probably poorly biased,
and are usually flagged due to elevated noise levels during
pre-processing (Section~\ref{sec:flagbadbol}).

This example illustrates some of the types of correlated signals and
patterns that PCA can identify in \scuba\ data. While there are
typically one or two strong signal components detected, in general,
the details can vary significantly from data set to data set, and the
lengths of the time-series analysed. The reason for this is probably
due to the fact that many of the noise sources are not stationary in
time (e.g., scan-synchronous noise which obviously depends on the scan
pattern, and the tuning of the subarrays). It should also be clear
from this example that while PCA offers some helpful insight into the
various sources of noise, it does not necessarily identify clean
patterns that can easily be modelled. For example, while there are
clearly correlation patterns along columns (Components~6 and 8--10,
particularly at 450\,\micron), as one might expect given the common
amplification chain, the intensities of these high-frequency signals
appear random in the eigenvalue maps, and a common-mode signal
estimated from the data for each column would not remove it. Finally,
we note that most of the high-frequency ($\gsim2$\,Hz) correlated
features that have been identified tend to be restricted to a small
number of detectors, and can often be identified and removed using
other (faster) means. What remains is predominantly low-frequency
noise. This conclusion is central to our data reduction strategy
described in the next section.

@@

The approach taken by SMURF to reducing \scuba\ data is to model (and
remove) predominantly low-frequency noise sources that are correlated
amongst detectors, and to iterate this process along with estimates of
the map. Significant experimentation with different models for noise
sources during \scuba\ commissioning required a highly flexible
software framework, and a configurable user interface. To achieve this
goal, while minimizing development time, it was decided to build SMURF
as a Starlink package \citep{2009ASPC..411..418J}, which provides
access to a large suite of libraries (including a commanding and
messaging interface, astrometric coordinate conversions and generation
of standard WCS information, file formats etc.). Furthermore,
Starlink\footnote{\url{http://www.starlink.ac.uk}} is
open-source\footnote{\url{https://github.com/Starlink}}, and already
used extensively at the Joint Astronomy Centre (host of the JCMT) for
many other systems \citep{jenness2011}, which helps with
interoperability. Though originally written in FORTRAN, many of the
core Starlink libraries are now ported to native C, or at least have a
C interface. It was therefore decided to develop SMURF in C as well
(rather than C++, for example, which would have required adding
further dependencies to Starlink), although we have taken an
object-oriented approach. For example, all of the data for a given
noise model are encapsulated in a C structure, and there is a standard
interface for all functions that handle models (member data and
functions for the class, respectively). In this way it is easy to
extend SMURF with new models. Parallelisation is incorporated in the
most time-consuming low-level routines using threads (e.g., performing
Fast Fourier Transforms, re-gridding the data), usually handling
either independent blocks of bolometers, or blocks of time, in each
thread, depending on the nature of the calculation. We have found that
for typical data sets the processing time scales well with the number
of central processing unit (CPU) cores, although beyond 8 the returns
are diminished due to tasks that cannot be parallelised (e.g., reading
the data from disk).

@@

First, the raw data are passed through a pre-processing stage which
corrects some of the more significant glitches, applies flat-field
corrections etc. Next, the iterative process begins (dashed box). Most
of the low-frequency noise is removed using common-mode subtraction
(\model{COM,GAI}, Section~\ref{sec:comgai}). Then, the extinction
correction (\model{EXT}, Section~\ref{sec:ext}) is applied. At this
point the data resemble the grey bolometer traces in
Fig.~\ref{fig:bolos_mix}, with some residual baseline drifts still
visible. These drifts are removed using a high-pass filter
(\model{FLT}, Section~\ref{sec:flt}) implemented with Fast Fourier
Transforms (FFTs). The residual signal is then re-gridded to estimate
the map. Finally, the map is used to estimate and remove the
astronomical signal contamination in the bolometer data (\model{AST},
Section~\ref{sec:ast}), leaving a relatively clean data set in which
to measure the white noise level of each bolometer (\model{NOI},
Section~\ref{sec:noi}). Since the common-mode and filtering stages
will have introduced ringing in the map in the vicinity of bright
sources, the entire process is iterated. Each component in the dashed
box of Fig.~\ref{fig:dimm} is re-calculated in sequence. In this way,
the second time the common-mode is calculated, for example, most of
the bright astronomical sources in the data have already been removed
in the previous iteration when the map was estimated, reducing the
amount of ringing in the map once it is re-estimated. Note that the
extinction correction is a multiplicative factor, and must be inverted
prior to re-calculating any of the additive model components to
preserve the data amplitude. Additive model components from the
previous iteration are generically added back into the time-series
immediately prior to their re-calculation. However, for the special
case of the common-mode and high-pass filter stages, they are replaced
simultaneously at the start of the iteration to assist with
convergence (see Sections~\ref{sec:converge}).

@@

The method used by SMURF is to average together multiple samples from
the original time-series, $x_i$ to estimate the lower-frequency output
time-series, $y_j$. In general there will not be an integer number of
samples from $x_i$ in each of the $y_j$, so fractional samples from
$x_i$ are used at the $y_j$ time boundaries. The algorithm is fast,
since each sample in $x_i$ need only be visited once. From a spectral
point of view, this boxcar average is equivalent to applying a
sinc-function low-pass filter. Such behaviour is desirable, since it
serves as an anti-aliasing filter; most non-white features above the
Nyquist frequency in $y_j$ that could contaminate the output data are
removed.

@@

In addition to replacing bad samples before the FFT, it is also
necessary to ensure that the data values at the start and end of the
time-series are similar. Since an FFT treats the data as a single
cycle in an infinitely repeating waveform, any large difference
between starting and ending values will effectively introduce sudden
steps at the start and end of each cycle, causing unwanted
oscillations (ringing) in the transform. Another consequence of the
cyclic nature of the FFT is that features at one end of the time
series can affect the filtered values at the other end of the time
series. Two methods are available to avoid these two problems:

@@

\item Apodisation: a number of samples at the start and end of each
bolometer time-series are multiplied by a cosine function in order to
roll the data values off smoothly to zero. The default number of
samples modified at each end of the time-series is given by half the
ratio of the sampling frequency to the lowest retained frequency. In
addition, each end of the time-series is padded with double this
number of zeros. This method is illustrated in Fig.~\ref{fig:pad2}. It
is not used by default as it reduces the amount of data available for
the map, and can significantly hinder very short observations (e.g.,
of calibrators when focussing the telescope).

@@

Despite the cleaning operations described in the previous sections,
the data from a given bolometer may be unusable due to it being poorly
biased, having an incorrect flatfield correction applied, or having
some other particularly pathological noise contamination. Most of
these bolometers can be flagged simply by identifying outliers in the
distribution of bolometer white noise levels. By default, SMURF
measures the PSD of all bolometers between 2 and 10\,Hz, after all
other pre-processing steps have been run (identical to the measurement
in Section~\ref{sec:noi}). Despite having significant low-frequency
noise (and possibly bright astronomical source) contamination, this
higher-frequency portion of the PSD is generally quite
clean. Furthermore, even if there is contamination, the purpose of
this measurement is to identify outliers, rather than provide a
meaningful absolute measurement of a given bolometer's white noise
level. Both high and low (e.g., due to an incorrect and very small
flatfield correction being applied) outliers from the centre of the
logarithm (to reduce the impact of outliers) of the bolometer noise
distribution are flagged.

@@

We then express the noise, $\mathbf{n}_i(t)$, as the sum of several
components,
%
\begin{equation}
  \mathbf{n}_i(t) = \mathbf{n}^\mathrm{w}_i(t) +
  g_i\mathbf{n}^\mathrm{c}(t) + \mathbf{n}^\mathrm{f}_i(t),
\label{eq:noise}
\end{equation}
%
where $\mathbf{n}^\mathrm{w}_i(t)$ is uncorrelated white noise,
$\mathbf{n}^\mathrm{c}(t)$ is a correlated or common-mode signal (with
an optional scale factor $g_i$ for each bolometer), and
$\mathbf{n}^\mathrm{f}_i(t)$ is (predominantly low-frequency) noise in
excess of the white noise level, that is either uncorrelated from
bolometer-to-bolometer, or at least does not have a simple correlation
relationship that would lead to it being included in
$\mathbf{n}^\mathrm{c}(t)$.

@@

\begin{table}
  \caption{Summary of the model components that can be fit to \scuba\
    time-series data with SMURF. Only the first group of models are
    typically fit to the data (\model{COM}--\model{NOI}) in the
    indicated order. The remaining models (\model{DKS}--\model{PLN})
    are usually omitted, although they are available as options.}
  \vspace{0.2cm}
  \centering
  \begin{tabular}{c|l}
    \hline
    Model & Description \\
    \hline
    \model{COM} & remove common-mode signal \\
    \model{GAI} & common-mode scaled to each bolometer \\
    \model{EXT} & extinction correction \\
    \model{FLT} & Fourier transform filter \\
    \model{AST} & map estimate of astronomical signal \\
    \model{NOI} & noise estimation \\
    \hline
    \model{DKS} & dark squid cleaning along columns \\
    \model{PLN} & 2-dimensional time-varying plane removal \\
    \model{SMO} & time-domain smoothing filter \\
    \model{TMP} & pointing as baseline template \\
    \hline
    \end{tabular}
  \label{tab:components}
\end{table}

@@

\begin{figure}
\centering
\includegraphics[width=\linewidth]{com.pdf}
\caption{A selection of typical bolometer time-series after
flatfielding and removal of a constant baseline. It can be seen that
most bolometers exhibit a common time variation overlaid with other
features. Empty squares indicate the locations of broken bolometers.}
\label{fig:com}
\end{figure}

The \model{COM} model is the common-mode signal itself
[$\mathbf{n}^\mathrm{c}(t)$ in Eq.~\ref{eq:noise}]. It is a single
time-series that is estimated by finding the mean of all bolometers at
each time step. Bolometer values that have been flagged as unusable
are excluded from the mean.

@@

It can be seen from Fig.~\ref{fig:com} that some bolometers depart
radically from the common-mode, indicating some problem with the
bolometer. Such bolometers are identified by calculating the Pearson
correlation coefficient between each bolometer time-series and the
common mode. Bolometers for which the correlation coefficient is below
a specified limit, or which have unusually high or low gains compared
to the other bolometers, are flagged as bad in order to omit them from
the final map. If the option described above for handling time-varying
sensitivities is used, then a correlation coefficient can be
determined for each individual block of time slices.  This allows
individual bad blocks to be rejected from a bolometer time-series,
rather than rejecting the whole bolometer. As a warning, rejecting
data based on outlier gain values can be misleading in cases where the
data are dominated by magnetic field pickup. For example, the
common-mode signal for the data in Fig.~\ref{fig:magpickup} would
resemble the azimuthal scan pattern, but clearly the gains will have
opposite signs (such that one or the other bolometer would be
erroneously flagged).

@@

Any extended astronomical emission on a scale comparable to or larger
than the spatial extent of the area used to estimate the common-mode
will contribute a similar signal to all bolometers. Therefore such
extended emission is indistinguishable from the other sources of
common-mode signal (e.g., atmosphere variations) and will be removed
by the \model{COM} model. This places a limit on the spatial extent of
astronomical structure that can be recovered.

@@

This model takes the FFT of the bolometer time-series data, and can
apply both high- and low-pass filters, as well as notch filters, at
hard frequency edges specified by the user. Alternatively, the
frequency edges of the filters may be defined in terms of an angular
scale, but converted into a frequency through knowledge of the mean
telescope slew speed. The time-series are generally gap-filled
(Section~\ref{sec:gaps}) before the transform to avoid ringing
(primarily caused by wrap-around discontinuities at the ends of the
time-series). Finally, a whitening filter may also be applied in which
a simple form, $1/f^\alpha + \mathrm{constant}$, is fit to the power
spectrum of each bolometer, and then the bolometer FFT is multiplied
by the square root of its inverse (though normalised to preserve the
white noise level).  The signals that are removed from the time-series
by this process are stored in the \model{FLT} container
array. Typically this model is used purely as a high-pass filter to
remove most of the residual $1/f$ noise following common-mode removal
[$\mathbf{n}^\mathrm{f}_i(t)$ in Eq.~\ref{eq:noise}]. Low-pass
filtering is redundant for two reasons: (i) SMURF already low-pass
filters and re-samples to a lower sample rate, as described in
Section~\ref{sec:downsamp}; and (ii) the act of re-gridding the data
to produce map estimates effectively low-pass filters the data below a
frequency that corresponds to the inverse of the crossing time of a
single map pixel. Notch filters have not been proven to be useful with
\scuba\ data, particularly since line features tend to move around
(i.e., a dynamic line-detection system would have to be developed, and
care would have to be taken that astronomical sources are not
suppressed).

@@

In addition to the signal map, a variance map $\mathbf{v}(x_j,y_j)$ is
estimated. The default procedure is to estimate this quantity given
the scatter in the weighted samples. This is accomplished by dividing
the biased weighted sample variance by the number of samples that went
into the average (akin to the formula for standard error on the mean,
but accounting for weights),
%
\begin{equation}
\label{eq:varmap}
\mathbf{v}(x_j,y_j) = \frac{\sum_j \mathbf{w}_j
                            \sum_j \mathbf{w}_j \mathbf{b}_j -
                            \left( \sum_j \mathbf{w}_j \mathbf{b}_j \right)^2 }
                           { N \left( \sum_j \mathbf{w}_j \right)^2 },
\end{equation}
%
where $N_j$ is the total number of bolometer samples that land in the
pixel. As written, this algorithm is numerically unstable if the two
terms in the numerator are large, and of nearly the same value:
floating point precision errors can cause the difference to be
significantly incorrect. In practice we use the more numerically
stable ``weighted incremental algorithm'' that calculates incremental
differences, as described in \citet{west1979}.

We decided not to use an unbiased estimator (e.g., the extension of
the common $(1/N-1)\sum_j (\mathbf{b}_j - \bar{\mathbf{b}})^2$
estimator using weights), since in practice it would require
accumulating an additional array of values at every map pixel, and
only results in a small difference where there are less than $\sim10$
samples per pixel (a situation that is almost never encountered in a
\scuba\ map, except in the edge pixels).
